---
title: "p8105_hw2_jer2243"
author: "Juliet Rowe"
date: "2023-09-27"
output: github_document
---

# Problem 1

## Load tidyverse
```{r}
library(tidyverse)
library(dplyr)
```


## Import and Clean pols-month Data

```{r clean_538_pols}
month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )

pols = 
  read_csv("fivethirtyeight_datasets/pols-month.csv") |>
  separate(mon, into = c("year", "month_num", "day"), convert = TRUE) |>
  mutate(
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) |>
  left_join(x = _, y = month_df) |> 
  select(year, month, everything(), -day, -starts_with("prez")) 
```

## Import and clean snp dataset

```{r}
snp = 
  read_csv("fivethirtyeight_datasets/snp.csv") |>
  separate(date, into = c("month", "day", "year"), convert = TRUE) |>
  arrange(year, month) |>
  mutate(month = month.name[month]) |>
  select(year, month, close) 
```


## Import and clean the unemployment dataset

```{r}
unemployment = 
  read_csv("fivethirtyeight_datasets/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, unemployment)
```


## Joining datasets
```{r merge_538}
data_538 = 
  left_join(pols, snp) |>
  left_join(x = _, y = unemployment)

str(data_538)
```

The `pols` data has `r nrow(pols)` observations and `r ncol(pols)` variables and tells us about the party affiliation distribution (democrat or republican) for governors and senators for a given year from years `r range(pols$year)[1]` to `r range(pols$year)[2]`. It also tells us whether the sitting president was a democrat or republican. The `snp` data has `r nrow(snp)` observations and `r ncol(snp)` variables, ranging from years `r range(snp$year)[1]` to `r range(snp$year)[2]`. The `unemployment` data has `r nrow(unemployment)` observations and `r ncol(unemployment)` variables ranging from years `r range(unemployment$year)[1]` to `r range(unemployment$year)[2]`. In Januarys in or after 1975 in which a democrat was president, the **average unemployment rate was `r filter(data_538, month == "January", year >= 1975, president == "dem") |> pull(unemployment) |> mean() |> round(2)`**.  The average unemployment rate over the same time period in which a republican was president was `r filter(data_538, month == "January", year >= 1975, president == "gop") |> pull(unemployment) |> mean() |> round(2)`.

# Problem 2

## Import and clean Mr. Trash Wheel Data

```{r}
library(readxl)
mr_trashwheel_data = read_excel("~/Desktop/p8105_hw2_jer2243/202309 Trash Wheel Collection Data.xlsx", sheet=1, range="A2:M586") |>
  janitor::clean_names()

mr_trashwheel_data_tidy = mr_trashwheel_data |>
  mutate(homes_powered = (weight_tons*500/30),
         name = "Mr. Trash Wheel",
         year = as.numeric(year))
  
```

## Import and clean Professor Trash Wheel Data

```{r}
library(readxl)
prof_trashwheel_data = read_excel("~/Desktop/p8105_hw2_jer2243/202309 Trash Wheel Collection Data.xlsx", sheet=2, range="A2:L108") |>
  janitor::clean_names()

prof_trashwheel_data_final = prof_trashwheel_data |>
  mutate(homes_powered = (weight_tons*500/30),
         name = "Professor Trash Wheel")


```

## Import and clean Gwynnda data

```{r}
library(readxl)
gwynnda_data = read_excel("~/Desktop/p8105_hw2_jer2243/202309 Trash Wheel Collection Data.xlsx", sheet=4, range="A2:K157") |>
  janitor::clean_names()

gwynnda_data_final = gwynnda_data |>
  mutate(homes_powered = (weight_tons*500/30),
         name="Gwynnda")
```

## Join the 3 trashwheel datasets
```{r}
trashwheel_data =
  bind_rows(mr_trashwheel_data_tidy, prof_trashwheel_data_final, gwynnda_data_final) |>
  janitor::clean_names()

```

The `trashwheel_data` data has `r nrow(trashwheel_data)` observations and `r ncol(trashwheel_data)` variables and describes the trash collected by different water wheels in the harbor of Baltimore, MD. The key variables are `weight_tons`, `homes_powered`, `dumpster`, and `month`, `year`, and `date`. `cigarette_butts` and `plastic_bottles` were two of the most common trash items picked up by the trashwheels, among other less common ones such as `sports balls` and `plastic_bags`.
Total weight of trash collected by Prof Trashwheel is `r sum(pull(prof_trashwheel_data, weight_tons))`. Total number of cigarette butts collected by Gwynnda in 2021 is `r sum(pull(filter(gwynnda_data, month == "July", year == "2021"), cigarette_butts))`


# Problem 3

## Import and clean MCI baseline data
```{r}
MCI_baseline = read_csv("data_mci/mci_baseline.csv", skip = 1) |>
  janitor::clean_names()

MCI_baseline_final = MCI_baseline |>
  filter(age_at_onset >= current_age | age_at_onset == ".") |>
  mutate(
    sex = as.character(sex),
    sex = case_match(sex,
                     "0" ~ "Female",
                     "1" ~ "Male"),
    apoe4 = as.character(apoe4),
    apoe4 = case_match(apoe4,
                       "0" ~ "Non-carrier",
                       "1" ~ "Carrier"),
    id = as.character(id))


```

The `MCI_baseline_final` data has `r nrow(MCI_baseline_final)` observations and `r ncol(MCI_baseline_final)` variables. This means that `r nrow(MCI_baseline_final)` participants were recruited (and fit the criteria of not having MCI at baseline).Of these participants, `r nrow(subset(MCI_baseline_final, age_at_onset != "."))` developed MCI.
The average age at the study baseline is `r mean(pull(MCI_baseline_final, current_age))`. 


The proportion of women in the study who are APOE4 carriers is `r sum(with(MCI_baseline_final, sex == "Female" | apoe4 == "Carrier")) / nrow(subset(MCI_baseline_final, sex == "Female")) * 100`%


## Import and clean MCI amyloid data
```{r}
MCI_amyloid = read_csv("data_mci/mci_amyloid.csv", skip=1) |>
  janitor::clean_names()

MCI_amyloid_final = MCI_amyloid |>
  rename(id=study_id) |>
  mutate(id = as.character(id))

```
The `MCI_amyloid_final` data has `r nrow(MCI_amyloid_final)` observations and `r ncol(MCI_amyloid_final)` variables. 

Some participants only appear in the cleaned amyloid dataset and not the cleaned baseline dataset, as the amyloid dataset has `r nrow(MCI_amyloid_final)` observations and the baseline only has `r nrow(MCI_baseline_final)`. 


## Join baseline and amyloid data and export
```{r}
joined_data = 
  inner_join(MCI_baseline_final, MCI_amyloid_final, by = "id")
write.csv(joined_data, file = "data_mci/joined_data.csv")
```

The final tidied dataset combining baseline and amyloid data only had full data for `r nrow(joined_data)` participants. The average age at baseline for this group was `r round(mean(pull(joined_data, current_age)), digits=2)` years. Of these participants, `r nrow(filter(joined_data, sex=="Female"))` were women.
The proportion of participants who were carriers of the APOE4 variant was `r sum(with(joined_data, apoe4 == "Carrier")) / nrow(joined_data) * 100`%.

